Supporting Information: Thermal sensitivity across forest vertical profiles: patterns, mechanisms, and ecological implications

### 2022-04-01 vertical-thermal-review.Rmd
further work on Tleaf section

Notes_S1<- "Notes S1"

@@

$T_{leaf}$ is strongly tied to $T_{air}$, and is further influenced by other micrometeorological drivers (section 1) and by leaf traits and stomatal conductance (see sections 3-4). Leaves adjust to their environment to approach, when possible, an optimal $T_{leaf}$ for photosynthesis and metabolism [@michaletzPlantThermoregulationEnergetics2015a <!--@perezPhotosyntheticHeatTolerances2020; @drakeNoEvidenceHomeostatic2020-->]. 
Leaves are often cooler than the air on clear nights due to radiative coupling with the very cold sky, and under some daytime conditions [cloudy skies, high wind speeds, and high $T_{air}$, @vogelLeavesLowestHighest2009; @rey-sanchez_spatial_2016;  @cavaleriColdbloodedForestsWarming2020].
Leaves can be warmer than air in full sunlight, especially under slow wind speeds and low $T_{air}$ [@doughtyAreTropicalForests2008].
$T_{leaf}$ is usually within a few degrees of $T_{air}$ [@cavaleriColdbloodedForestsWarming2020].

@@

### 2.1 Vertical gradients in leaf temperature

Biophysical principles (`r Notes_S1`) determine how elevated $T_{leaf}$ is above $T_{air}$ across vertical gradients in closed-canopy forests.
High radiation in the upper canopy (Fig. `r fig_NEON_vertical`) implies that upper canopy leaves often have higher $T_{leaf}-T_{air}$, and thus are often warmer than understory leaves (Fig. `r fig_leaf_T`).
However, higher wind speeds (Fig. `r fig_NEON_vertical`) reduce leaf-air temperature differences [Fig. `r fig_leaf_T`b, @niinemetsPhotosyntheticAcclimationSimultaneous2004; @bonan_ecological_2016]. 
In addition, adaptive leaf traits that increase $g_b$ and $g_s$ (see sections 3, 4.1) mediate the direct effects of solar radiation on upper canopy leaves and result in higher rates of heat loss (Fig. `r fig_leaf_T`b-e). 
In contrast, in the lower canopy, greater RH and lower wind speeds may limit $\lambda E$ (through reduced VPD and $g_b$), increasing $T_{leaf}$ [Fig. `r fig_leaf_T`, @perezIncreasingHumidityThreatens2018; @tibbittsHumidityPlants1979].
Leaves in these normally-shaded locations experience rapid light and temperature surges during sunflecks (>10$^\circ$C)  [@waySunflecksTreesForests2012].

@@

`r Notes_S1`

>>>>

Notes_leafTbiophysics<- "S1"

@@

$T_{leaf}$ is strongly tied, and usually within a few degrees of, $T_{air}$ [@cavaleriColdbloodedForestsWarming2020], but can be substantially warmer or cooler under certain conditions (Fig. `r fig_leaf_T`; Note `r Notes_leafTbiophysics`). 
Deviation of $T_{leaf}$ from $T_{air}$ is influenced by other micrometeorological drivers and by leaf traits and stomatal conductance (Note `r Notes_leafTbiophysics`), all of which vary across forest vertical gradients (Fig. `r fig_schematic`).
Leaves adjust to their environment to approach, when possible, an optimal $T_{leaf}$ for photosynthesis and metabolism [@michaletzPlantThermoregulationEnergetics2015a <!--@perezPhotosyntheticHeatTolerances2020; @drakeNoEvidenceHomeostatic2020-->].

@@

Biophysical principles (Note `r Notes_leafTbiophysics`) shape $T_{leaf}-T_{air}$ across vertical gradients in closed-canopy forests.
High radiation in the upper canopy (Fig. `r fig_NEON_vertical`) implies that upper canopy leaves often have higher $T_{leaf}-T_{air}$, and thus are often warmer than understory leaves (Fig. `r fig_leaf_T`).
However, higher wind speeds (Fig. `r fig_NEON_vertical`) reduce leaf-air temperature differences [Fig. `r fig_leaf_T`b, @niinemetsPhotosyntheticAcclimationSimultaneous2004; @bonan_ecological_2016]. 
In addition, adaptive leaf traits that increase $g_b$ and $g_s$ (see sections 3, 4.1) mediate the direct effects of solar radiation on upper canopy leaves and result in higher rates of heat loss (Fig. `r fig_leaf_T`b-e). 
In contrast, in the lower canopy, greater RH and lower wind speeds may limit $\lambda E$ (through reduced VPD and $g_b$), increasing $T_{leaf}$ [Fig. `r fig_leaf_T`, @perezIncreasingHumidityThreatens2018; @tibbittsHumidityPlants1979].
Leaves in these normally-shaded locations experience rapid light and temperature surges during sunflecks (>10$^\circ$C)  [@waySunflecksTreesForests2012].

@@

`r Notes_leafTbiophysics`

### 2022-04-01 vertical-thermal-review_SI.Rmd
further work on Tleaf section

>>>>

Leaves can be substantially warmer or cooler under certain conditions. 
They can be  warmer than air in full sunlight, especially under slow wind speeds and low $T_{air}$ [@doughtyAreTropicalForests2008].
Leaves are often cooler than the air on clear nights due to radiative coupling with the very cold sky, and under some daytime conditions [cloudy skies, high wind speeds, and high $T_{air}$, @vogelLeavesLowestHighest2009; @rey-sanchez_spatial_2016;  @cavaleriColdbloodedForestsWarming2020].
